---
title: "Get EPA Pollutant Values at the Census Block Levels"
author: "Nisha Narayanan (narayanannisha5@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: flatly
    highlight: pygments
---
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PARGASITE Overview

**Pollution-Associated Risk Geospatial Analysis SITE**

PARGASITE uses publicly available data from the United States Environmental Protection Agency (EPA). We have no affiliation with the EPA. From this data, we generated monthly and yearly raster files (Jan 1997 to Dec 2021) for PM2.5, Ozone, NO2, SO2, and CO covering the contiguous US using inverse distance weighted interpolation from the 5 nearest EPA monitoring stations. The R package allows the user to obtain a more customized output at the Census block level compared to the website (http://pargasite.org/) and work with the raster layers directly. The package functions uses the raster layers built from EPA data and overlays them on the shapefiles provided by the US Census (see the section Tigris to learn how to fetch the shape files) to extract the corresponding pollutant values at the MMSA, County, Tract and ZIPCODE levels. The final output has the mean, median, and standard deviation computed for each grouped value for every level.

You can refer to our published work here:
Greenblatt RE, Himes BE. Facilitating Inclusion of Geocoded Pollution Data into Health Studies. AMIA Jt Summits Transl Sci Proc. 2019;2019:553–561.(PMID: [31259010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6568125/))

## Installing PARGASITE

Install the prerequisite R packages if they do not exist:

```{r eval=F, echo=T}
library(devtools)
devtools::install_github("HimesGroup/pargasite", subdir = "pkg")
```

Load necessary packages: 

```{r eval=T, echo=T,warning=FALSE,message=FALSE}
library(pargasite)
library(pander)
```

## Downloading Census Shape Files from Tigris

The inbuilt functions of the PARGASITE package requires the user to supply the shape files of the census blocks required during the function call. These can be extracted from the Tigris package. Details about specific function is discussed in subsequent sections.

```{r eval = F, echo =T}
install.packages('tigris') #install package if not done previously
```

Load the package
```{r}
library(tigris)
```

## Extracting Pollutant Data at the MMSA Level

The first step is to get the MMSA shape files from tigris and it can be done using the "core_based_statistical_areas" function. Check the column names of the final data frame. If the column names are not GEOID for the GEOID and NAME for the MMSA names, rename the columns as required. Followed by this, use the "getMMSALevelPollutantValue" from PARGASITE to get the pollutant value for each MMSA. The function requires three variables to be passed along - year (anything between 1997 to 2021), Pollutant (one among PM2.5, CO, Ozone, NO2, and SO2), and the MMSA shape file data frame (as provided by tigris).The final data frame returned will have 5 columns (GEOID, MMSA Names, Mean Pollutant Value, Median Pollutant Value, SD of Pollutant values). See example below.

```{r, eval = F, echo = T}
mmsa_shp <- core_based_statistical_areas(cb = FALSE, resolution = "500k", year = "2021") #replace 2021 with the required year

mmsa_level_pollutant_value <- getMMSALevelPollutantValue(2021, "PM2.5", mmsa_shp) # replace 2021 with required year
```

## Extracting Pollutant Data at the Counties Level

The first step is to get the counties shape files from tigris and it can be done using the "counties" function. Check the column names of the final data frame. If the column names are not GEOID for the GEOID and NAMELSAD for the county names (ends with county), rename the columns as required. Followed by this, use the "getCountyLevelPollutantValue" from PARGASITE to get the pollutant value for each county. The function requires three variables to be passed along - year (anything between 1997 to 2021), Pollutant (one among PM2.5, CO, Ozone, NO2, and SO2), and the counties shape file data frame (as provided by tigris). The final data frame returned will have 5 columns (GEOID, County Names, Mean Pollutant Value, Median Pollutant Value, SD of Pollutant values). See example below.

```{r, eval = F, echo = T}
county_shp <- counties_shp <- counties(state = NULL, cb = FALSE, resolution = "500k", year = "2021") #replace 2021 with the required year

county_level_pollutant_value <- getCountyLevelPollutantValue(2021, "PM2.5", county_shp) # replace 2021 with required year
```

## Extracting Pollutant Data at the Census Tract Level

The first step is to get the counties shape files from tigris and it can be done using the "tracts" function. Check the column names of the final data frame. If the column names are not GEOID for the GEOID and NAMELSAD for the tracts names (ends with county), rename the columns as required. Followed by this, use the "getTractLevelPollutantValue" from PARGASITE to get the pollutant value for each county. The function requires three variables to be passed along - year (anything between 1997 to 2021), Pollutant (one among PM2.5, CO, Ozone, NO2, and SO2), and the tracts shape file data frame (as provided by tigris). The final data frame returned will have 5 columns (GEOID, Census Tract Names, Mean Pollutant Value, Median Pollutant Value, SD of Pollutant values). See example below.

```{r, eval = F, echo = T}
tract_shp <- tracts(state = NULL, county = NULL, cb = TRUE, year = "2021") #replace 2021 with the required year

tract_level_pollutant_value <- getTractLevelPollutantValue(2021, "PM2.5", tract_shp) # replace 2021 with required year

```

## Extracting Pollutant Data at the ZIPCODE Level

The first step is to get the counties shape files from tigris and it can be done using the "zctas" function. Check the column names of the final data frame. If the column name is not ZCTA5CE20 for the ZIPCODE, rename the columns as required (GEOID in this case is same as the ZIPCODE - rename either of those before using in the function). Followed by this, use the "getZIPLevelPollutantValue" from PARGASITE to get the pollutant value for each county. The function requires three variables to be passed along - year (anything between 1997 to 2021), Pollutant (one among PM2.5, CO, Ozone, NO2, and SO2), and the ZIPCODE shape file data frame (as provided by tigris). The final data frame returned will have 4 columns (ZIPCODE, Mean Pollutant Value, Median Pollutant Value, SD of Pollutant values). See example below.

```{r, eval = F, echo = T}
zipcode_shp <- zipcode_master <- zctas(cb = FALSE, starts_with = NULL, year = NULL, state = "2021") #replace 2021 with the required year

zipcode_level_pollutant_value <- getZIPLevelPollutantValue(2021, "PM2.5", zipcode_shp) # replace 2021 with required year
```

___

#### Session information

```{r sessioninfo, eval=T, echo=F}
pander(sessionInfo())
```

